---
title: Truck Report
author:
  - name: A Bender
    email: alice@example.com
    affiliation: Khoury
abstract: |
  This is the report on the analysis of truck factor in GitHub repositories. 
output: rticles::acm_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(readr)
library(rticles)

```

# Introduction

Our goal is to compute for each file the Degree of Authorship (DOA) measure:

```
DOA(f) = 3.293+1.098*FA+0.164*DL-0.321*log(1+AC)
```

where `f` is file; `FA` is true if the author is the first author of `f`; `DL` is the number of commits made by the author; `AC` is the total number of commits.

Then for each file, the author with highest DOA will be consider the owner of that file.

Then for each project, the truck factor (TF) is the number of owners you have to remove to affect half of the files in the project.


```{r}
strip <- function(name) strsplit(name,"[.]csv[.]gz")

all_meta <- list.files("data/commit_metadata")
name <-strip(all_meta[45])

m <-  read_csv( paste0("data/commit_metadata/", name ,".csv.gz"))
print(paste0("read ", nrow(m), " rows from project ", name ))
f <-  read_csv( paste0("data/commit_files/",name,".csv.gz"))
print(paste0("read ", nrow(f), " rows from project ",name))

m %>%
  rename(author=`author name`, time =`author timestamp`) %>%
  select(author, time, hash) -> m1

f %>%
  rename(add = `added lines`) %>%
  select(hash, add, file) %>%
  filter(add != 0) %>%
  filter( grepl( "\\.java$|\\.js$|\\.R$|\\.c$" , file)) -> f1

print(paste0("removed ", nrow(f)-nrow(f1), " from f (", round(100*(nrow(f)-nrow(f1))/nrow(f),1),"%)"))

m1 %>%
  left_join(f1, by="hash") %>%
  filter ( ! is.na(add)) %>%
  select(author, time, file)  -> joined

joined %>%
  group_by(file) %>%
  arrange(time, .by_group=TRUE) %>%
  summarize(first=first(author)) ->first_authors

joined %>%
  group_by(file, author) %>%
  summarize(my_commits=n()) -> commits_per_author_file

joined %>%
  group_by(file) %>%
  summarize(all_commits=n()) -> commits_per_file

first_authors %>%
  left_join(commits_per_author_file, by = "file") %>%
  left_join(commits_per_file, by = "file") %>%
  rename(DL=my_commits,AC=all_commits) %>%
  mutate(FA= author == first) -> df

DOA <- 3.293 +(1.098*df$FA)+0.164*df$DL-0.321*log(1+df$AC)

df %>% 
  mutate(doa=DOA)  %>%
  select(file,author,doa) -> df2

df2 %>%
  group_by(file) %>%
  arrange(desc(doa)) %>%
  summarize(owner=first(author)) -> owners

file_count <- length(unique(df2$file))
  
owners %>%
  group_by(owner) %>%
  summarize(n=n(), perc = n/file_count) %>%
  arrange(desc(perc)) %>%
  mutate(cs = cumsum(perc), half = cs < .5, tf = cumsum(half) + 1) ->  result


```

# Methods

How are we going to do it

# Results

What we got


